% Glossary of acronyms and key terms (distinct from the References list).
% Linked from the first emergence of each acronym via \gloss{...}.

\begin{onehalfspacing}

\chapter*{Glossary}
\label{ch:glossary}
\hypertarget{glossary}{}
\addcontentsline{toc}{chapter}{Glossary}

This glossary collects the acronyms and domain-specific terms used in the thesis, each with a short description. Where an acronym first appears in the text it is hyperlinked to this list. The glossary is separate from the bibliographic references at the end of the document.

\section*{Acronyms}

\renewcommand{\arraystretch}{1.2}
\small
\begin{longtable}{@{}l >{\raggedright\arraybackslash\hyphenpenalty=10000\exhyphenpenalty=10000}p{3.5cm} >{\raggedright\arraybackslash\hyphenpenalty=10000\exhyphenpenalty=10000}p{6.6cm}@{}}
\toprule
\textbf{Acronym} & \textbf{Expansion} & \textbf{Description} \\
\midrule
\endfirsthead
\toprule
\textbf{Acronym} & \textbf{Expansion} & \textbf{Description} \\
\midrule
\endhead
AI    & Artificial Intelligence            & Computational methods that learn or reason from data. \\
ADF   & Augmented Dickey--Fuller           & Statistical test for whether a series is stationary (has no unit root). \\
ATR   & Average True Range                 & A measure of recent price volatility, used to size stops and targets. \\
AUC   & Area Under the ROC Curve           & Ranking quality of a classifier; $0.5$ is random, $1.0$ is perfect. \\
B\&H  & Buy-and-Hold                       & Passive benchmark that buys the asset once and holds it. \\
BTC   & Bitcoin                            & The cryptocurrency studied throughout the thesis. \\
CMF   & Chaikin Money Flow                 & Volume-weighted momentum oscillator. \\
EMA   & Exponential Moving Average         & Moving average that weights recent prices more heavily. \\
ETH   & Ethereum                           & Second-largest cryptocurrency; used for the ETH/BTC cross. \\
GRU   & Gated Recurrent Unit               & A recurrent-network cell; prototyped but not retained. \\
LGBM  & LightGBM                           & Gradient-boosted decision-tree framework; the tabular agent. \\
LOB   & Limit Order Book                   & The full set of resting buy and sell orders at each price. \\
LSTM  & Long Short-Term Memory             & A recurrent-network cell with gating for long sequences. \\
MACD  & Moving Average Convergence Divergence & Momentum indicator from the gap between a fast and slow EMA. \\
MAS   & Multi-Agent System                 & A set of autonomous agents combined into one decision-maker. \\
MaxDD & Maximum Drawdown                   & Largest peak-to-trough loss of an equity~curve. \\
MLP   & Multi-Layer Perceptron             & A basic feed-forward neural network. \\
MPS   & Metal Performance Shaders          & Apple's GPU backend, used for local model training. \\
OHLCV & Open, High, Low, Close, Volume     & The five fields of a price candle. \\
OOS   & Out-Of-Sample                      & Data held out from training, used for honest evaluation. \\
PPO   & Proximal Policy Optimisation       & A reinforcement-learning algorithm; prototyped, not retained. \\
RL    & Reinforcement Learning             & Learning to act by maximising cumulative reward. \\
RSI   & Relative Strength Index            & Momentum oscillator bounded between $0$ and $100$. \\
SMA   & Simple Moving Average              & Unweighted average of recent prices. \\
SSM   & State-Space Model                  & Sequence model with a learned latent state (the Mamba agent). \\
TBM   & Triple Barrier Method              & Labelling that classifies a trade by which of three barriers it hits first. \\
TCN   & Temporal Convolutional Network     & Sequence model built from dilated causal convolutions. \\
TFI   & Taker Flow Imbalance               & Net aggressive buy-versus-sell pressure from taker volume. \\
VWAP  & Volume-Weighted Average Price      & Average traded price weighted by~volume. \\
WFO   & Walk-Forward Optimisation          & Repeatedly train on the past and test on the next period. \\
\bottomrule
\end{longtable}

\section*{Key terms}

\renewcommand{\arraystretch}{1.2}
\small
\begin{longtable}{@{}l >{\raggedright\arraybackslash\hyphenpenalty=10000\exhyphenpenalty=10000}p{10.2cm}@{}}
\toprule
\textbf{Term} & \textbf{Description} \\
\midrule
\endfirsthead
\toprule
\textbf{Term} & \textbf{Description} \\
\midrule
\endhead
Heterogeneous & Components built from categorically different learning paradigms, each with its own bias and failure mode. \\
Hybrid & The integration of dissimilar components into a single decision-making system. \\
Fund-of-agents & The framing of the system as a portfolio of autonomous sub-strategies combined by a capital allocator. \\
Inverse-volatility risk parity & Allocation rule that weights each agent inversely to its trailing return volatility, equalising risk contributions. \\
Random-bracket null & Significance test that re-runs an agent's signal as random permutations through the same bracket, to separate timing skill from risk-management effects. \\
Meta-labelling & A two-stage scheme in which a second model predicts whether a primary trading signal will be profitable. \\
Fractional differentiation & A transform that makes a price series stationary while keeping long-memory content, set by a fractional order $d$. \\
Sharpe ratio & Mean excess return divided by its standard deviation, annualised; the main risk-adjusted metric. \\
Sortino ratio & A Sharpe variant that penalises only downside volatility. \\
Alpha & Return in excess of a passive benchmark over the same period. \\
Regime & A persistent market condition (here: chop, bull, or bear) with characteristic trend and volatility. \\
\bottomrule
\end{longtable}

\end{onehalfspacing}
